---
title: "- RNA-seq - DE Analysis (3) -"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

<style>
body{text-align: justify}
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

*File creation: February, 22nd 2019*  
*Update: January, 15th 2020*   

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(data.table)
library(Biobase)
library(limma)
library(edgeR)
library(here)
library(doMC)
library(DT)
library(ComplexHeatmap)
```
```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE}
load(here("output", "2019-02-22_RNA-seq_DE_3.RData"))
# save.image(here("output", "2019-02-22_RNA-seq_DE_3.RData"))
```

# Statistical Analysis
______________________

## Differential analysis btw outcomes within each time point

Statistical analyses are performed w/ the `limma` R package (well-established package for RNA-seq and microarray analysis). A linear model is fitted to each gene, and empirical Bayes moderated t-statistics are used to assess differences in expression. Within each time point & PD-1+TIGIT+, four contrasts of interest are investigated   

* NR vs R - PD-1-TIGIT-;  

* NR vs R - PD-1+TIGIT+;  

* NR vs R - PD-1+;  

* NR vs R - TIGIT+.   

We subset the whole dataset into different subsets depending on the combination time point & treatment. An absolute log2-fold change cutoff of 1 and a false discovery rate (FDR) cutoff of 5% are used to determine differentially expressed genes (DEGs); whereas a false discovery rate (FDR) cutoff of 5% is used to determine differentially expressed gene sets (GSEA).   

### Time point T0 - DEGs

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- DE Analysis
countData <- readRDS(file = here("output", "RNA_count.rds"))
#- Subsetting
countData.tmp <- countData[, which(countData$time.point == "T0")]
#- phenoData
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
phenoData.tmp %>% View("pData")
#- Normalization factor
norm.tmp <- calcNormFactors(countData.tmp)
#- Design matrix
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
#- Contrast matrix
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
#- Voom transformation
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)
normData_1 <- voomData.tmp$E
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Fitting
fit1.tmp <- lmFit(object = voomData.tmp, design = myDesign.tmp)
fit2.tmp <- contrasts.fit(fit = fit1.tmp, contrasts = aovCon.tmp)
fit2.tmp <- eBayes(fit = fit2.tmp, trend = FALSE)

registerDoMC(2)
results_DEGs_1 <- foreach(i = 1:ncol(aovCon.tmp)) %dopar%
{
  results.tmp <- topTable(fit = fit2.tmp, adjust.method = "fdr", coef = i, number = nrow(voomData.tmp), sort = "P")
  results.tmp <- data.table(Gene = rownames(results.tmp), results.tmp)
  write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("DEGs_T0_", colnames(aovCon.tmp)[i], ".csv")))
  results.tmp[, Direction := ifelse(adj.P.Val < 0.05 & sign(logFC) == 1 & abs(logFC) >= 1, "Up",
                                    ifelse(adj.P.Val < 0.05 & sign(logFC) == -1 & abs(logFC) >= 1, "Down", "NotDE"))]
  return(results.tmp)
}
names(results_DEGs_1) <- colnames(aovCon.tmp)

results_DEGs_1$NR_vs_R_DNEG[Direction != "NotDE"] # 0 DEG
results_DEGs_1$NR_vs_R_DPOS[Direction != "NotDE"] # 1 DEG
results_DEGs_1$NR_vs_R_PD1[Direction != "NotDE"] # 0 DEG
results_DEGs_1$NR_vs_R_TIGIT[Direction != "NotDE"] # 0 DEG
```

There are (FDR 5% & log2-FC > 1)  

* NR vs R - PD-1-TIGIT- - __0 DEG__;  

* NR vs R - PD-1+TIGIT+ - __1 DEG__;  

* NR vs R - PD-1+ - __0 DEG__;  

* NR vs R - TIGIT+ - __0 DEG__.   

**Volcano plots**

```{r volcano_1, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=3}
###--- Volcano Plot
data_ggplot <- bind_rows(data.table(bind_rows(results_DEGs_1), ID = c(rep(names(results_DEGs_1)[1], 975), rep(names(results_DEGs_1)[2], 975), rep(names(results_DEGs_1)[3], 975), rep(names(results_DEGs_1)[4], 975))))

data_ggplot %>%
  mutate(ID = str_replace(ID, pattern = "_vs_", replacement = " vs ")) %>%
  mutate(ID = str_replace(ID, pattern = "_", replacement = " - ")) %>%
  mutate(ID = factor(ID, levels = c("NR vs R - DNEG", "DNEG vs PD1", "NR vs R - DPOS", "NR vs R - PD1", "NR vs R - TIGIT"))) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) +
    geom_point(size = .5, alpha = .5) + 
    geom_vline(xintercept = c(-1, 1), color = "grey20", lty = 3) + 
    geom_hline(yintercept = c(-log10(0.05)), color = "grey20", lty = 3) +
    scale_color_manual(guide = FALSE, values = c("Down" = "royalblue", "Up" = "indianred3", "NotDE" = "grey20")) +
    facet_wrap(~ ID, ncol = 5) +
    labs(x = "logFC", y = "-log10(FDR)") +
    theme_bw()
```

**Tables**

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, results="asis"}
###--- Top genes
DT_print <- list()
for(i in 1:4)
{
  results.tmp <- results_DEGs_1[[i]][, c("Gene", "logFC", "P.Value", "adj.P.Val")]
  colnames(results.tmp) <- c("Gene", "logFC", "Pvalue", "FDR")
  DT_print[[i]] <- results.tmp
}
contrast <- paste0("DEGs - ", c("NR vs R - DNEG",
                                "NR vs R - DPOS", 
                                "NR vs R - PD1", 
                                "NR vs R - TIGIT"))
datatable(DT_print[[1]], caption = contrast[1], options = list(pageLength = 5))
datatable(DT_print[[2]], caption = contrast[2], options = list(pageLength = 5))
datatable(DT_print[[3]], caption = contrast[3], options = list(pageLength = 5))
datatable(DT_print[[4]], caption = contrast[4], options = list(pageLength = 5))
```

**Boxplot**

```{r boxplot_1, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=3, fig.height=4}
###--- Boxplot
countData <- readRDS(file = here("output", "RNA_count.rds"))
countData <- countData[, colnames(normData_1)]

data.table(CXCL13 = normData_1["CXCL13", ], 
           pData(countData)) %>%
  filter(fraction.desc == "PD-1+TIGIT+") %>%
  filter(time.point == "T0") %>%
  ggplot(aes(x = outcome, y = CXCL13)) +
    geom_boxplot(aes(color = outcome), outlier.shape = NULL) +
    geom_point(aes(color = patient.id)) +
    labs(y = "CXCLR3 - normalized gene expression") +
    theme_classic() +
    theme(legend.position = "none", axis.title.x = element_blank())
```

### Time point T0 - GSEA

Three databases are used (from http://software.broadinstitute.org/gsea/msigdb/collections.jsp)  

* KEGG pathways;  

* Hallmark pathways;  

* Immunologic signatures - c7.  

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- GSEA
gene.sets <- readRDS(file = here("data", "gene-sets", "genesets_human.rds"))
countData <- readRDS(file = here("output", "RNA_count.rds"))

#- countData & norm
countData.tmp <- countData[, which(countData$time.point == "T0")]
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
norm.tmp <- calcNormFactors(countData.tmp)
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)

#- Get indices
registerDoMC(2)
indices.list <- foreach(i = 1:length(gene.sets)) %dopar%
{
  indices.tmp <- limma::ids2indices(gene.sets[[i]], rownames(normData_1))
  indices.tmp <- indices.tmp[sapply(indices.tmp, length) >= 5]
  return(indices.tmp)
}
names(indices.list) <- names(gene.sets)

#- GSEA - camera
registerDoMC(2)
results.GSEA_1 <- foreach(i = 1:length(results_DEGs_1)) %dopar%
{
  GSEA.tmp <- foreach(j = 1:length(indices.list)) %do%
  {
    results.tmp <- camera(voomData.tmp, indices.list[[j]], design = myDesign.tmp, contrast = aovCon.tmp[, i], sort = TRUE)
    results.tmp <- data.table(`Gene set` = rownames(results.tmp), results.tmp)
    results.tmp[, Genes := paste(rownames(voomData.tmp)[unlist(indices.list[[j]][`Gene set`])], collapse = ", "), by = `Gene set`]
    write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("GSEA_T0_", names(indices.list)[j], "_", colnames(aovCon.tmp)[i], ".csv")))
    results.tmp <- results.tmp[FDR < 0.05]
    return(results.tmp)
  }
  names(GSEA.tmp) <- names(indices.list)
  return(GSEA.tmp)
}
names(results.GSEA_1) <- names(results_DEGs_1)
```

There are (FDR 5%)  

* NR vs R - PD-1-TIGIT- - __1 gene set__ - KEGG;  

* NR vs R - PD-1-TIGIT- - __1 gene set__ - Hallmark;  

* NR vs R - PD-1-TIGIT- - __0 gene set__ - c7;  

* NR vs R - PD-1+TIGIT+ - __0 gene set__ - KEGG;  

* NR vs R - PD-1+TIGIT+ - __1 gene set__ - Hallmark;  

* NR vs R - PD-1+TIGIT+ - __0 gene set__ - c7;  

* NR vs R - PD-1+ - __0 gene set__ - KEGG;  

* NR vs R - PD-1+ - __1 gene set__ - Hallmark;  

* NR vs R - PD-1+ - __0 gene set__ - c7;  

* NR vs R - TIGIT+ - __1 gene set__ - KEGG;     

* NR vs R - TIGIT+ - __1 gene set__ - Hallmark;     

* NR vs R - TIGIT+ - __0 gene set__ - c7.  

Please look at files (results) in the **output/output_2019-02-22/** folder.  

### Time point M1 - DEGs

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- DE Analysis
countData <- readRDS(file = here("output", "RNA_count.rds"))
#- Subsetting
countData.tmp <- countData[, which(countData$time.point == "M1")]
#- phenoData
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
phenoData.tmp %>% View("pData")
#- Normalization factor
norm.tmp <- calcNormFactors(countData.tmp)
#- Design matrix
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
#- Contrast matrix
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
#- Voom transformation
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)
normData_2 <- voomData.tmp$E
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Fitting
fit1.tmp <- lmFit(object = voomData.tmp, design = myDesign.tmp)
fit2.tmp <- contrasts.fit(fit = fit1.tmp, contrasts = aovCon.tmp)
fit2.tmp <- eBayes(fit = fit2.tmp, trend = FALSE)

registerDoMC(2)
results_DEGs_2 <- foreach(i = 1:ncol(aovCon.tmp)) %dopar%
{
  results.tmp <- topTable(fit = fit2.tmp, adjust.method = "fdr", coef = i, number = nrow(voomData.tmp), sort = "P")
  results.tmp <- data.table(Gene = rownames(results.tmp), results.tmp)
  write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("DEGs_M1_", colnames(aovCon.tmp)[i], ".csv")))
  results.tmp[, Direction := ifelse(adj.P.Val < 0.05 & sign(logFC) == 1 & abs(logFC) >= 1, "Up",
                                    ifelse(adj.P.Val < 0.05 & sign(logFC) == -1 & abs(logFC) >= 1, "Down", "NotDE"))]
  return(results.tmp)
}
names(results_DEGs_2) <- colnames(aovCon.tmp)

results_DEGs_2$NR_vs_R_DNEG[Direction != "NotDE"] # 0 DEG
results_DEGs_2$NR_vs_R_DPOS[Direction != "NotDE"] # 0 DEG
results_DEGs_2$NR_vs_R_PD1[Direction != "NotDE"] # 0 DEG
results_DEGs_2$NR_vs_R_TIGIT[Direction != "NotDE"] # 0 DEG
```

There are (FDR 5% & log2-FC > 1)  

* NR vs R - PD-1-TIGIT- - __0 DEG__;  

* NR vs R - PD-1+TIGIT+ - __0 DEG__;  

* NR vs R - PD-1+ - __0 DEG__;  

* NR vs R - TIGIT+ - __0 DEG__.   

**Volcano plots**

```{r volcano_2, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=3}
###--- Volcano Plot
data_ggplot <- bind_rows(data.table(bind_rows(results_DEGs_2), ID = c(rep(names(results_DEGs_2)[1], 975), rep(names(results_DEGs_2)[2], 975), rep(names(results_DEGs_2)[3], 975), rep(names(results_DEGs_2)[4], 975))))

data_ggplot %>%
  mutate(ID = str_replace(ID, pattern = "_vs_", replacement = " vs ")) %>%
  mutate(ID = str_replace(ID, pattern = "_", replacement = " - ")) %>%
  mutate(ID = factor(ID, levels = c("NR vs R - DNEG", "DNEG vs PD1", "NR vs R - DPOS", "NR vs R - PD1", "NR vs R - TIGIT"))) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) +
    geom_point(size = .5, alpha = .5) + 
    geom_vline(xintercept = c(-1, 1), color = "grey20", lty = 3) + 
    geom_hline(yintercept = c(-log10(0.05)), color = "grey20", lty = 3) +
    scale_color_manual(guide = FALSE, values = c("Down" = "royalblue", "Up" = "indianred3", "NotDE" = "grey20")) +
    facet_wrap(~ ID, ncol = 5) +
    labs(x = "logFC", y = "-log10(FDR)") +
    theme_bw()
```

**Tables**

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, results="asis"}
###--- Top genes
DT_print <- list()
for(i in 1:4)
{
  results.tmp <- results_DEGs_2[[i]][, c("Gene", "logFC", "P.Value", "adj.P.Val")]
  colnames(results.tmp) <- c("Gene", "logFC", "Pvalue", "FDR")
  DT_print[[i]] <- results.tmp
}
contrast <- paste0("DEGs - ", c("NR vs R - DNEG",
                                "NR vs R - DPOS", 
                                "NR vs R - PD1", 
                                "NR vs R - TIGIT"))
datatable(DT_print[[1]], caption = contrast[1], options = list(pageLength = 5))
datatable(DT_print[[2]], caption = contrast[2], options = list(pageLength = 5))
datatable(DT_print[[3]], caption = contrast[3], options = list(pageLength = 5))
datatable(DT_print[[4]], caption = contrast[4], options = list(pageLength = 5))
```

### Time point M1 - GSEA

Three databases are used (from http://software.broadinstitute.org/gsea/msigdb/collections.jsp)  

* KEGG pathways;  

* Hallmark pathways;  

* Immunologic signatures - c7.  

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- GSEA
gene.sets <- readRDS(file = here("data", "gene-sets", "genesets_human.rds"))
countData <- readRDS(file = here("output", "RNA_count.rds"))

#- countData & norm
countData.tmp <- countData[, which(countData$time.point == "M1")]
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
norm.tmp <- calcNormFactors(countData.tmp)
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)

#- Get indices
registerDoMC(2)
indices.list <- foreach(i = 1:length(gene.sets)) %dopar%
{
  indices.tmp <- limma::ids2indices(gene.sets[[i]], rownames(normData_2))
  indices.tmp <- indices.tmp[sapply(indices.tmp, length) >= 5]
  return(indices.tmp)
}
names(indices.list) <- names(gene.sets)

#- GSEA - camera
registerDoMC(2)
results.GSEA_2 <- foreach(i = 1:length(results_DEGs_2)) %dopar%
{
  GSEA.tmp <- foreach(j = 1:length(indices.list)) %do%
  {
    results.tmp <- camera(voomData.tmp, indices.list[[j]], design = myDesign.tmp, contrast = aovCon.tmp[, i], sort = TRUE)
    results.tmp <- data.table(`Gene set` = rownames(results.tmp), results.tmp)
    results.tmp[, Genes := paste(rownames(voomData.tmp)[unlist(indices.list[[j]][`Gene set`])], collapse = ", "), by = `Gene set`]
    write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("GSEA_M1_", names(indices.list)[j], "_", colnames(aovCon.tmp)[i], ".csv")))
    results.tmp <- results.tmp[FDR < 0.05]
    return(results.tmp)
  }
  names(GSEA.tmp) <- names(indices.list)
  return(GSEA.tmp)
}
names(results.GSEA_2) <- names(results_DEGs_2)
```

There are (FDR 5%)  

* NR vs R - PD-1-TIGIT- - __0 gene set__ - KEGG;  

* NR vs R - PD-1-TIGIT- - __3 gene sets__ - Hallmark;  

* NR vs R - PD-1-TIGIT- - __14 gene sets__ - c7;  

* NR vs R - PD-1+TIGIT+ - __1 gene set__ - KEGG;  

* NR vs R - PD-1+TIGIT+ - __3 gene sets__ - Hallmark;  

* NR vs R - PD-1+TIGIT+ - __41 gene sets__ - c7;  

* NR vs R - PD-1+ - __4 gene sets__ - KEGG;  

* NR vs R - PD-1+ - __0 gene set__ - Hallmark;  

* NR vs R - PD-1+ - __0 gene set__ - c7;  

* NR vs R - TIGIT+ - __1 gene set__ - KEGG;     

* NR vs R - TIGIT+ - __0 gene set__ - Hallmark;     

* NR vs R - TIGIT+ - __3 gene sets__ - c7.    

Please look at files (results) in the **output/output_2019-02-22/** folder.  

### Time point M2 - DEGs

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- DE Analysis
countData <- readRDS(file = here("output", "RNA_count.rds"))
#- Subsetting
countData.tmp <- countData[, which(countData$time.point == "M2")]
#- phenoData
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
phenoData.tmp %>% View("pData")
#- Normalization factor
norm.tmp <- calcNormFactors(countData.tmp)
#- Design matrix
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
#- Contrast matrix
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
#- Voom transformation
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)
normData_3 <- voomData.tmp$E
```
```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- Fitting
fit1.tmp <- lmFit(object = voomData.tmp, design = myDesign.tmp)
fit2.tmp <- contrasts.fit(fit = fit1.tmp, contrasts = aovCon.tmp)
fit2.tmp <- eBayes(fit = fit2.tmp, trend = FALSE)

registerDoMC(2)
results_DEGs_3 <- foreach(i = 1:ncol(aovCon.tmp)) %dopar%
{
  results.tmp <- topTable(fit = fit2.tmp, adjust.method = "fdr", coef = i, number = nrow(voomData.tmp), sort = "P")
  results.tmp <- data.table(Gene = rownames(results.tmp), results.tmp)
  write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("DEGs_M2_", colnames(aovCon.tmp)[i], ".csv")))
  results.tmp[, Direction := ifelse(adj.P.Val < 0.05 & sign(logFC) == 1 & abs(logFC) >= 1, "Up",
                                    ifelse(adj.P.Val < 0.05 & sign(logFC) == -1 & abs(logFC) >= 1, "Down", "NotDE"))]
  return(results.tmp)
}
names(results_DEGs_3) <- colnames(aovCon.tmp)

results_DEGs_3$NR_vs_R_DNEG[Direction != "NotDE"] # 0 DEG
results_DEGs_3$NR_vs_R_DPOS[Direction != "NotDE"] # 0 DEG
results_DEGs_3$NR_vs_R_PD1[Direction != "NotDE"] # 0 DEG
results_DEGs_3$NR_vs_R_TIGIT[Direction != "NotDE"] # 0 DEG
```

There are (FDR 5% & log2-FC > 1)  

* NR vs R - PD-1-TIGIT- - __0 DEG__;  

* NR vs R - PD-1+TIGIT+ - __0 DEG__;  

* NR vs R - PD-1+ - __0 DEG__;  

* NR vs R - TIGIT+ - __0 DEG__.   

**Volcano plots**

```{r volcano_3, echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, fig.width=12, fig.height=3}
###--- Volcano Plot
data_ggplot <- bind_rows(data.table(bind_rows(results_DEGs_3), ID = c(rep(names(results_DEGs_3)[1], 975), rep(names(results_DEGs_3)[2], 975), rep(names(results_DEGs_3)[3], 975), rep(names(results_DEGs_3)[4], 975))))

data_ggplot %>%
  mutate(ID = str_replace(ID, pattern = "_vs_", replacement = " vs ")) %>%
  mutate(ID = str_replace(ID, pattern = "_", replacement = " - ")) %>%
  mutate(ID = factor(ID, levels = c("NR vs R - DNEG", "DNEG vs PD1", "NR vs R - DPOS", "NR vs R - PD1", "NR vs R - TIGIT"))) %>%
  ggplot(aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) +
    geom_point(size = .5, alpha = .5) + 
    geom_vline(xintercept = c(-1, 1), color = "grey20", lty = 3) + 
    geom_hline(yintercept = c(-log10(0.05)), color = "grey20", lty = 3) +
    scale_color_manual(guide = FALSE, values = c("Down" = "royalblue", "Up" = "indianred3", "NotDE" = "grey20")) +
    facet_wrap(~ ID, ncol = 5) +
    labs(x = "logFC", y = "-log10(FDR)") +
    theme_bw()
```

**Tables**

```{r echo=FALSE, eval=TRUE, message=FALSE, warning=FALSE, results="asis"}
###--- Top genes
DT_print <- list()
for(i in 1:4)
{
  results.tmp <- results_DEGs_3[[i]][, c("Gene", "logFC", "P.Value", "adj.P.Val")]
  colnames(results.tmp) <- c("Gene", "logFC", "Pvalue", "FDR")
  DT_print[[i]] <- results.tmp
}
contrast <- paste0("DEGs - ", c("NR vs R - DNEG",
                                "NR vs R - DPOS", 
                                "NR vs R - PD1", 
                                "NR vs R - TIGIT"))
datatable(DT_print[[1]], caption = contrast[1], options = list(pageLength = 5))
datatable(DT_print[[2]], caption = contrast[2], options = list(pageLength = 5))
datatable(DT_print[[3]], caption = contrast[3], options = list(pageLength = 5))
datatable(DT_print[[4]], caption = contrast[4], options = list(pageLength = 5))
```

### Time point M2 - GSEA

Three databases are used (from http://software.broadinstitute.org/gsea/msigdb/collections.jsp)  

* KEGG pathways;  

* Hallmark pathways;  

* Immunologic signatures - c7.  

```{r echo=TRUE, eval=FALSE, message=FALSE, warning=FALSE}
###--- GSEA
gene.sets <- readRDS(file = here("data", "gene-sets", "genesets_human.rds"))
countData <- readRDS(file = here("output", "RNA_count.rds"))

#- countData & norm
countData.tmp <- countData[, which(countData$time.point == "M2")]
phenoData.tmp <- pData(countData.tmp)
phenoData.tmp$group <- paste(phenoData.tmp$fraction, phenoData.tmp$outcome, sep = "_")
norm.tmp <- calcNormFactors(countData.tmp)
myDesign.tmp <- model.matrix(~ 0 + group, data = phenoData.tmp)
colnames(myDesign.tmp) <- str_remove(string = colnames(myDesign.tmp), pattern = "group")
aovCon.tmp <- makeContrasts(NR_vs_R_DNEG = (DNEG_NR - DNEG_R),
                            NR_vs_R_DPOS = (DPOS_NR - DPOS_R),
                            NR_vs_R_PD1 = (PD1_NR - PD1_R),
                            NR_vs_R_TIGIT = (TIGIT_NR - TIGIT_R),
                            levels = myDesign.tmp)
voomData.tmp <- voom(counts = countData.tmp, design = myDesign.tmp, lib.size = colSums(exprs(countData.tmp)) * norm.tmp)

#- Get indices
registerDoMC(2)
indices.list <- foreach(i = 1:length(gene.sets)) %dopar%
{
  indices.tmp <- limma::ids2indices(gene.sets[[i]], rownames(normData_3))
  indices.tmp <- indices.tmp[sapply(indices.tmp, length) >= 5]
  return(indices.tmp)
}
names(indices.list) <- names(gene.sets)

#- GSEA - camera
registerDoMC(2)
results.GSEA_3 <- foreach(i = 1:length(results_DEGs_3)) %dopar%
{
  GSEA.tmp <- foreach(j = 1:length(indices.list)) %do%
  {
    results.tmp <- camera(voomData.tmp, indices.list[[j]], design = myDesign.tmp, contrast = aovCon.tmp[, i], sort = TRUE)
    results.tmp <- data.table(`Gene set` = rownames(results.tmp), results.tmp)
    results.tmp[, Genes := paste(rownames(voomData.tmp)[unlist(indices.list[[j]][`Gene set`])], collapse = ", "), by = `Gene set`]
    write_csv(x = results.tmp, path = here("output", "output_2019-02-22/", paste0("GSEA_M2_", names(indices.list)[j], "_", colnames(aovCon.tmp)[i], ".csv")))
    results.tmp <- results.tmp[FDR < 0.05]
    return(results.tmp)
  }
  names(GSEA.tmp) <- names(indices.list)
  return(GSEA.tmp)
}
names(results.GSEA_3) <- names(results_DEGs_3)
```

There are (FDR 5%)  

* NR vs R - PD-1-TIGIT- - __5 gene sets__ - KEGG;  

* NR vs R - PD-1-TIGIT- - __3 gene sets__ - Hallmark;  

* NR vs R - PD-1-TIGIT- - __0 gene set__ - c7;  

* NR vs R - PD-1+TIGIT+ - __4 gene sets__ - KEGG;  

* NR vs R - PD-1+TIGIT+ - __3 gene sets__ - Hallmark;  

* NR vs R - PD-1+TIGIT+ - __0 gene set__ - c7;  

* NR vs R - PD-1+ - __4 gene sets__ - KEGG;  

* NR vs R - PD-1+ - __2 gene sets__ - Hallmark;  

* NR vs R - PD-1+ - __0 gene sets__ - c7;  

* NR vs R - TIGIT+ - __4 gene sets__ - KEGG;     

* NR vs R - TIGIT+ - __1 gene set__ - Hallmark;     

* NR vs R - TIGIT+ - __4 gene sets__ - c7.  

Please look at files (results) in the **output/output_2019-02-22/** folder.  

